---
title: "Concord, MA 1951-2020 (plants and birds)"
author: "Karen Ornelas"
date: "2024-03-05"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

load tidyverse
```{r}
#install.packages("tidyverse")
library(tidyverse)
```

read csv file 
```{r}
Concord_data <- read.csv("Concord_phenology_dataset.csv")
```

Remove all unnescary columns before turning the data into long format 
```{r}
Concord_data <- Concord_data [-c(4,5,6,7,8)]
```


Turn data into long format
```{r}
# change all data into interegers
Concord_data <- Concord_data %>%
  mutate(across(X1845:X2020, as.numeric))

Concord_long <- Concord_data %>% 
  pivot_longer(
    cols = 5:125,
    names_to = "year",
    values_to = "DOY")
```

Take out the X from the year 
```{r}
Concord_long$year <-  
  str_remove(
    Concord_long$year, "X")
```

remove na in duration and this will also take out our ice out
```{r}
Concord_long <- 
  Concord_long %>% 
    drop_na()
```


add column for lat and long 
```{r}
Concord_long <- Concord_long %>%
  mutate(Longitude = "-71.3333333")

Concord_long <- Concord_long %>%
  mutate(Latitude = "42.4500000")

```

```{r}
## Convert DOY to year month date
Concord_long <- Concord_long %>% mutate(date= as.Date(DOY-1, 
              origin=paste0(year,"-01-01"))) 
            
```

Create a new test database with only scientfic name to match TSN
```{r}
#install.packages("taxize")
library(taxize)

Concord_long$Scientific_name <- paste(Concord_long$Genus,Concord_long$Species)

Concord_long <-  Concord_long %>% relocate(Scientific_name,.after = Species)

tsn_df <- Concord_long %>%
  distinct(Scientific_name)

tsn_df <- tsn_df %>% 
  mutate(TSN = get_tsn(tsn_df$Scientific_name, rows = 1))

```

Take out the NAs from tsn_df 
```{r}
NA_TSN <- tsn_df[is.na(tsn_df$TSN), ] 
```

Use Fuzzy function in worldflora 
```{r}
#install.packages("WorldFlora")
library(WorldFlora)

#install.packages("fuzzyjoin")
library(fuzzyjoin)

options(timeout = max(1000, getOption("timeout")))

#WFO.download()

WFO.remember("classification.csv")

fuzzy_match <- WFO.match.fuzzyjoin(spec.data = NA_TSN$Scientific_name, WFO.data = WFO.data, fuzzydist.max = 2)

# have WFO condense the above dataframe down to one clear match per species
one_to_one <- WFO.one(fuzzy_match)

# save the output so we don't have to run those again because they take up wild amounts of RAM!
write_csv(one_to_one, "corrected_taxonomy.csv")

#replace scientificName with correct ITIS accepted names to be able to find the TSN
one_to_one$scientificName <- str_replace(one_to_one$scientificName,"Lysimachia arvensis", "Anagallis arvensis")
one_to_one$scientificName <-str_replace(one_to_one$scientificName,"Scorzoneroides autumnalis", "Leontodon autumnalis")
one_to_one$scientificName <-str_replace(one_to_one$scientificName,"Nemopanthes mucronata", "Ilex mucronata")
one_to_one$scientificName <-str_replace(one_to_one$scientificName,"Muraltia pauciflora", "
Polygala paucifolia")
one_to_one$scientificName <-str_replace(one_to_one$scientificName,"Ailanthus altissima var. altissima", "Ailanthus altissima")
one_to_one$scientificName <-str_replace(one_to_one$scientificName,"Athyrium filix-femina subsp. filix-femina", "Athyrium filix-femina")
one_to_one$scientificName <-str_replace(one_to_one$scientificName,"Homalosorus pycnocarpos", "Diplazium pycnocarpon")
one_to_one$scientificName <-str_replace(one_to_one$scientificName,"Botrychium virginianum subsp. virginianum", "Botrypus virginianus")
one_to_one$scientificName <-str_replace(one_to_one$scientificName,"Cystopteris fragilis subsp. fragilis", "Cystopteris fragilis")
one_to_one$scientificName <-str_replace(one_to_one$scientificName,"Dryopteris boottii", "Dryopteris X boottii")
one_to_one$scientificName <-str_replace(one_to_one$scientificName,"Onoclea pensylvatica", "Matteuccia struthiopteris var. pensylvanica")


# run the updated names through `taxize` to get updates TSN
one_to_one_TSN <- one_to_one %>% 
  mutate(TSN = get_tsn(one_to_one$scientificName, rows = 1))


# input TSN for remaining NA values
one_to_one_TSN <- one_to_one_TSN %>% 
  mutate(
    TSN = ifelse(spec.name == "Polygala pauciflora", "29306", TSN))

one_to_one_TSN <- one_to_one_TSN %>% 
  mutate(
    TSN = ifelse(spec.name == "Aspidium vari incisum", "504531", TSN))

one_to_one_TSN <- one_to_one_TSN %>% 
  mutate(
    TSN = ifelse(spec.name == "Rhamnus fragularia", "28579", TSN))

one_to_one_TSN <- one_to_one_TSN %>% 
  mutate(
    TSN = ifelse(spec.name == "Scutellaria laterifolia", "32765", TSN))



```

Rename the scientific_name column in one to one TSN to match tsn_df
```{r}
colnames(one_to_one_TSN)[colnames(one_to_one_TSN) == 'scientificName'] <- 'Scientific_name'
```

Change TSN from chr to numeric
Relocate positions of columns to match tsn_df
```{r}
tsn_df <- tsn_df %>% 
  mutate(TSN = as.numeric(tsn_df$TSN)) 

one_to_one_TSN <- one_to_one_TSN %>% 
  mutate(TSN = as.numeric(one_to_one_TSN$TSN)) 

one_to_one_TSN <-  one_to_one_TSN %>% 
  relocate(Scientific_name, .before = spec.name.ORIG)

one_to_one_TSN <-  one_to_one_TSN %>% 
  relocate(TSN, .after = Scientific_name)
```

Rename all misspelled Scientific_name in tsn_df to correct name located in one_to_one_TSN
Replace all NA values in tsn_df with values from one_to_one_TSN
```{r}
tsn_df <- tsn_df %>% 
  mutate(Scientific_name = str_squish(Scientific_name)) %>% 
  drop_na(Scientific_name)
one_to_one_TSN <- one_to_one_TSN %>% 
  mutate(spec.name = str_squish(spec.name),
         Scientific_name = str_squish(Scientific_name))

misspelled <- one_to_one_TSN$spec.name

for (i in 1:nrow(tsn_df)) {
  if (tsn_df$Scientific_name[i] %in% misspelled){
    index <- which(one_to_one_TSN$spec.name == tsn_df$Scientific_name [i])
    tsn_df$Scientific_name[i] <- one_to_one_TSN$Scientific_name[index]
  } else {
    tsn_df$Scientific_name[i] <- tsn_df$Scientific_name [i]
  }
}


### run this after top for loop is confirmed to work
for (i in 1:nrow(tsn_df)){
  
  if (is.na(tsn_df$TSN[i])){
    
    if (str_detect(tsn_df$Scientific_name[i], "sp.")) {
      index <- which(one_to_one_TSN$spec.name.ORIG == tsn_df$Scientific_name [i])
      tsn_df$TSN[i] <-  one_to_one_TSN$TSN[index]
    } else {
      index <- which(one_to_one_TSN$Scientific_name == tsn_df$Scientific_name [i])
      tsn_df$TSN[i] <-  one_to_one_TSN$TSN[index]
    }
    
  } else {
    tsn_df$TSN [i] == tsn_df$TSN [i]
  }
}
  


i=159

```

Fix all misspelled names in Concord_Long -- do we fix species and genus 
```{r}

Concord_corrected_name <- read.csv("corrected_taxonomy.csv")

Concord_long <- Concord_long %>% 
  mutate(Scientific_name = str_squish(Scientific_name),
         Scientific_name = str_trim(Scientific_name))
Concord_corrected_name <- Concord_corrected_name %>% 
  mutate(spec.name.ORIG = str_squish(spec.name.ORIG),
         ScientificName = str_squish(scientificName),
         spec.name.ORIG = str_trim(spec.name.ORIG),
         ScientificName = str_trim(scientificName))

misspelled <- Concord_corrected_name$spec.name
for (i in 1:nrow(Concord_long)) {
  if (str_trim(Concord_long$Scientific_name[i]) %in% misspelled){
    index <- which(Concord_corrected_name$spec.name.ORIG == str_trim(Concord_long$Scientific_name[i]))
    Concord_long$Scientific_name[i] <- Concord_corrected_name$scientificName[index]
  } else {
    Concord_long$Scientific_name[i] <- Concord_long$Scientific_name [i]
  }
}

```


Final step
```{r}
tsn_df <- tsn_df %>% distinct()

Clean_Concord_Data <- full_join(Concord_long, tsn_df)

write_csv(Clean_Concord_Data,"Clean_Thoreau_Data")

```


```



