---
title: "Concord, MA 1951-2020 (plants and birds)"
author: "Karen Ornelas"
date: "2024-03-05"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Concord, MA Phenology Data

### Set-Up

Install and load the package we will use.

```{r}
# Install remotes if not already installed
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}

# Install specific versions of packages
# remotes::install_version("tidyverse", version = "2.0.0")
# remotes::install_version("taxize", version = "0.9.99")
# remotes::install_version("fuzzyjoin", version = "0.1.6")
# remotes::install_version("WorldFlora", version = "1.14-5")

# Load those packages
packages <- c("tidyverse", "WorldFlora", "taxize", "fuzzyjoin")
lapply(packages, function(pkg) library(pkg, character.only = TRUE))
```

Bring in the dataset.

```{r}
### Read csv file
Concord_data <- read.csv("Concord_phenology_dataset.csv")
```

## Data Wrangling

```{r}
### Remove all unnecessary columns before turning the data into long format 
Concord_data <- Concord_data [-c(4,5,6,7,8)]
```

The data are currently in a wide format. We want them to be in a long format with one column for the year and one column for the day of year.

```{r}
### Turn data into long format (This will make our year its own column)
# change all data into integers rather than characters
#in `mutate()`: ℹ In argument: `across(X1845:X2020, as.numeric)`.Caused by warning:! NAs introduced by coercion --- ignore this warning, there are values in the DOY column that are -x- which  indicates a representative decadal value as opposed to an annual observation aka an average

Concord_data <- Concord_data %>%
  mutate(across(X1845:X2020, as.numeric))


Concord_long <- Concord_data %>% 
  pivot_longer(
    cols = starts_with("X"),
    names_to = "year",
    values_to = "DOY")
```

```{r}
### Take out the X from the year 
Concord_long$year <-  
  str_remove(
    Concord_long$year, "X")
```

```{r}
### Remove all NA in the data set Concord_long. 
Concord_long <- 
  Concord_long %>% 
    drop_na()
```

```{r}
### Add a column for latitude and longitude
Concord_long <- Concord_long %>%
  mutate(Longitude = "-71.3333333")

Concord_long <- Concord_long %>%
  mutate(Latitude = "42.4500000")

```

```{r}
## Convert DOY to year month date
Concord_long <- Concord_long %>% mutate(date= as.Date(DOY-1, 
              origin=paste0(year,"-01-01"))) 
            
```

## Adding TSNs

We want to find TSNs (taxonomic serial numbers) for all of the species. This helps address issues that arise from changes in plant taxonomies.

First, we need to create a column in the dataset that has both the Genus and the species together.

```{r}
### Create a new test database with only scientific name to then find TSN (Taxonomic Scientific Number) for each species

Concord_long$Scientific_name <- paste(Concord_long$Genus,Concord_long$Species)

Concord_long <-  Concord_long %>% relocate(Scientific_name,.after = Species)
```

```{r}
# some of the species names need to be replaced (ITIS?)
Concord_long$Scientific_name <- str_replace(Concord_long$Scientific_name,"Lysimachia arvensis", "Anagallis arvensis")
Concord_long$Scientific_name <-str_replace(Concord_long$Scientific_name,"Scorzoneroides autumnalis", "Leontodon autumnalis")
Concord_long$Scientific_name <-str_replace(Concord_long$Scientific_name,"Nemopanthes mucronata", "Ilex mucronata")
Concord_long$Scientific_name <-str_replace(Concord_long$Scientific_name,"Muraltia pauciflora", "
Polygala paucifolia")
Concord_long$Scientific_name <-str_replace(Concord_long$Scientific_name,"Ailanthus altissima var. altissima", "Ailanthus altissima")
Concord_long$Scientific_name <-str_replace(Concord_long$Scientific_name,"Athyrium filix-femina subsp. filix-femina", "Athyrium filix-femina")
Concord_long$Scientific_name <-str_replace(Concord_long$Scientific_name,"Homalosorus pycnocarpos", "Diplazium pycnocarpon")
Concord_long$Scientific_name <-str_replace(Concord_long$Scientific_name,"Botrychium virginianum subsp. virginianum", "Botrypus virginianus")
Concord_long$Scientific_name <-str_replace(Concord_long$Scientific_name,"Cystopteris fragilis subsp. fragilis", "Cystopteris fragilis")
Concord_long$Scientific_name <-str_replace(Concord_long$Scientific_name,"Dryopteris boottii", "Dryopteris X boottii")
Concord_long$Scientific_name <-str_replace(Concord_long$Scientific_name,"Onoclea pensylvatica", "Matteuccia struthiopteris var. pensylvanica")
Concord_long$Scientific_name <-str_replace(Concord_long$Scientific_name,"Aspidium vari incisum", "Polystichum acrostichoides")

## make it so the species for Rhamnus fragularia and Scutellaria laterifolia appear in the Scientfic_name column
Concord_long$Scientific_name[which(Concord_long$Species == "fragularia")] <- "Rhamnus fragularia"

Concord_long$Scientific_name[which(Concord_long$Species == "laterifolia")] <- "Scutellaria laterifolia" 
```

### Getting TSNs for Each Species

We use the `tsn_df` function from the `taxize` package to find the TSN values

```{r}
tsn_df <- Concord_long %>%
  distinct(Scientific_name)

tsn_df <- tsn_df %>% 
  mutate(TSN = get_tsn(tsn_df$Scientific_name, rows = 1))
write_csv(tsn_df, "data_intermediate/tsn_df.csv")
```

Some did not have TSN values found. Save them in a new dataframe.

```{r}
### Take out the NAs from tsn_df  
NA_TSN <- tsn_df[is.na(tsn_df$TSN), ] 
```

For the species that did not have TSN values, part of the reason might be because they are misspelled.

```{r}
### Use Fuzzy function in WorldFlora to find the corrected scientific name for any mispelled scientific names

# options(timeout = max(2000, getOption("timeout")))
# 
# WFO.download()
```

We are using `WorldFlora` and `fuzzyjoin` to find the names of species which are misspelled.

```{r}
# read in the taxonomy file from WFO
# if this .csv is not in the project directory, run the code chunk above
WFO.remember("classification.csv")

fuzzy_match <- WFO.match.fuzzyjoin(spec.data = NA_TSN$Scientific_name, WFO.data = WFO.data, fuzzydist.max = 2)
#write_csv(fuzzy_match, "data_intermediate/fuzzy_match.csv")

# Have WFO condense the above data frame down to one clear match per species
one_to_one <- WFO.one(fuzzy_match)
```

Some of the names just have to be corrected manually.

```{r}
#Replace scientificName with correct ITIS accepted names to be able to find the TSN

one_to_one <- one_to_one %>% 
  mutate(scientificName = case_when(scientificName == "Lysimachia arvensis" ~ "Anagallis arvensis",
                                    scientificName == "Scorzoneroides autumnalis" ~ "Leontodon autumnalis",
                                    scientificName == "Nemopanthes mucronata" ~ "Ilex mucronata",
                                    scientificName == "Muraltia pauciflora" ~ "Polygala paucifolia",
                                    scientificName == "Ailanthus altissima var. altissima" ~ "Ailanthus altissima",
                                    scientificName == "Athyrium filix-femina subsp. filix-femina" ~ "Athyrium filix-femina",
                                    scientificName == "Homalosorus pycnocarpos" ~ "Diplazium pycnocarpon",
                                    scientificName == "Botrychium virginianum subsp. virginianum" ~ "Botrypus virginianus",
                                    scientificName == "Cystopteris fragilis subsp. fragilis" ~ "Cystopteris fragilis",
                                    scientificName == "Dryopteris boottii" ~ "Dryopteris X boottii",
                                    scientificName == "Onoclea pensylvatica" ~ "Matteuccia struthiopteris var. pensylvanica",
                                    scientificName == "Raphanus raphanistrum subsp. sativus" ~ "Raphanus sativus",
                                    scientificName == "Ampelopsis glandulosa var. brevipedunculata" ~ "Ampelopsis brevipedunculata",
                                    scientificName == "Prenanthes virgatum" ~ "Prenanthes autumnalis",
                                    scientificName == "Nabalus autumnalis" ~ "Prenanthes autumnalis",
                                    TRUE ~ scientificName))
```

We now run the correctly spelled names of the species through the `get_tsn` function to get their TSNs.

```{r}
# Run the updated names through `taxize` to get updates TSN
one_to_one_TSN <- one_to_one %>% 
  mutate(TSN = get_tsn(one_to_one$scientificName, rows = 1))
write_csv(one_to_one_TSN, "data_intermediate/one_to_one_TSN.csv")
```

Again, some have to be changed manually.

```{r}
# one_to_one_TSN <- read_csv("data_intermediate/one_to_one_TSN.csv")

# Input TSN for remaining NA values
one_to_one_TSN <- one_to_one_TSN %>% 
  mutate(TSN = case_when(spec.name == "Polygala pauciflora" ~ "29306",
                         spec.name == "Aspidium vari incisum" ~ "17675",
                         spec.name == "Rhamnus fragularia" ~ "28579",
                         spec.name == "Scutellaria laterifolia" ~ "32765",
                         spec.name == "Raphanus sativus" ~ "23290"))
```

```{r}
## make it so the spec.name for Rhamnus fragularia and Scutellaria laterifolia appear in the scientificName column
one_to_one_TSN$scientificName[which(one_to_one_TSN$spec.name == "Rhamnus fragularia")] <- "Rhamnus fragularia"

one_to_one_TSN$scientificName[which(one_to_one_TSN$spec.name == "Scutellaria laterifolia")] <- "Scutellaria laterifolia"

one_to_one_TSN$scientificName[which(one_to_one_TSN$spec.name == "Raphanus sativus")] <- "Raphanus sativus"

one_to_one_TSN$scientificName[which(one_to_one_TSN$spec.name == "Tabacum longiflorum")] <- "Tabacum longiflorum"

one_to_one_TSN$scientificName[which(one_to_one_TSN$TSN == "17675")] <- "Polystichum acrostichoides"
```

### Merge TSN Datasets

We need to join the datasets together so that the Concord dataset had the TSNs associated with each species. This will take a number of steps because we also want to correct the taxonomic names that are out-of-date or misspelled.

```{r}
### Rename the scientific_name column in one to one TSN to match tsn_df to help with matching up TSN to the correct species
colnames(one_to_one_TSN)[colnames(one_to_one_TSN) == 'scientificName'] <- 'Scientific_name'
```

```{r}
### Change TSN from character to numeric 

tsn_df <- tsn_df %>% 
  mutate(TSN = as.numeric(TSN)) 

one_to_one_TSN <- one_to_one_TSN %>% 
  mutate(TSN = as.numeric(TSN)) 

### Relocate positions of columns to match tsn_df
one_to_one_TSN <-  one_to_one_TSN %>% 
  relocate(Scientific_name, .before = spec.name.ORIG)

one_to_one_TSN <-  one_to_one_TSN %>% 
  relocate(TSN, .after = Scientific_name)
```

First, we need to update the taxonomic names in the dataframe with all of the TSNs.

```{r}
### Rename all misspelled Scientific_name in tsn_df to correct name located in one_to_one_TSN

# remove whitespace to fix some matching issues
tsn_df <- tsn_df %>% 
  mutate(Scientific_name = str_squish(Scientific_name),
         Scientific_name = str_trim(Scientific_name))
one_to_one_TSN <- one_to_one_TSN %>% 
  mutate(spec.name = str_squish(spec.name),
         spec.name = str_trim(spec.name),
         Scientific_name = str_squish(Scientific_name),
         Scientific_name = str_trim(Scientific_name)) 

# replace misspelled names in the TSN dataframe with the corrected names
misspelled <- one_to_one_TSN$spec.name

for (i in 1:nrow(tsn_df)) {
  if (tsn_df$Scientific_name[i] %in% misspelled){
    index <- which(one_to_one_TSN$spec.name == tsn_df$Scientific_name [i])
    tsn_df$Scientific_name[i] <- one_to_one_TSN$Scientific_name[index]
  } else {
    tsn_df$Scientific_name[i] <- tsn_df$Scientific_name [i]
  }
}
```

Now that all of the names in the `tsn_df` dataframe are corrected, we can replace missing TSNs with the ones that we got from running the corrected names through the `get_tsn` function.

```{r}
## Replace all NA values in tsn_df with values from one_to_one_TSN



for (i in 1:nrow(tsn_df)){
  
  if (is.na(tsn_df$TSN[i])){
    
    if (str_detect(tsn_df$Scientific_name[i], "sp\\.") | str_detect(tsn_df$Scientific_name[i], "\\?")) {
      index <- which(one_to_one_TSN$spec.name.ORIG == tsn_df$Scientific_name [i])
      tsn_df$TSN[i] <-  one_to_one_TSN$TSN[index]
    } else {
      index <- which(one_to_one_TSN$Scientific_name == tsn_df$Scientific_name [i])
      tsn_df$TSN[i] <-  one_to_one_TSN$TSN[index]
    }
    
  } else {
    tsn_df$TSN [i] == tsn_df$TSN [i]
  }
}

write_csv(one_to_one_TSN, "data_intermediate/one_to_one_TSN.csv")
write_csv(tsn_df, "data_intermediate/tsn_df_corrected.csv")
```

## Correct Concord Species and Add TSNs

```{r}
# Write one_to_one_TSN to "corrected_taxonomy.csv"
# write_csv(one_to_one_TSN, "corrected_taxonomy.csv")
# OR
# Read in "corrected_taxonomy.csv"
# one_to_one_TSN <- read_csv("data_intermediate/one_to_one_TSN.csv")
```

Fix all of the misspelled taxonomic names in the Concord dataset by replacing them with the corrected name from the `one_to_one_TSN` dataframe

```{r}
### Str_squish and Str_Trim will help remove white space, double spaces between words, and make it recognizable by R to match up with the tsn_df
Concord_long <- Concord_long %>% 
  mutate(Scientific_name = str_squish(Scientific_name),
         Scientific_name = str_trim(Scientific_name))
one_to_one_TSN <- one_to_one_TSN %>% 
  mutate(spec.name.ORIG = str_squish(spec.name.ORIG),
         ScientificName = str_squish(Scientific_name),
         spec.name.ORIG = str_trim(spec.name.ORIG),
         ScientificName = str_trim(Scientific_name))


### Fix all misspelled names in Concord_Long using the names within the mispelled dataframe

misspelled <- one_to_one_TSN$spec.name

for (i in 1:nrow(Concord_long)) {
  if (str_trim(Concord_long$Scientific_name[i]) %in% misspelled){
    index <- which(one_to_one_TSN$spec.name.ORIG == str_trim(Concord_long$Scientific_name[i]))
    Concord_long$Scientific_name[i] <- one_to_one_TSN$Scientific_name[index]
  } else {
    Concord_long$Scientific_name[i] <- Concord_long$Scientific_name [i]
  }
}

```

```{r}
### Make sure there is one tsn for only one scientific_name
tsn_df <- tsn_df %>% distinct()

### Full join will match up the scientific names in both dataframes to fill in the missing TSN in Concord_long
Clean_Concord_Data <- full_join(Concord_long, tsn_df)

### .CSV that includes the final clean dataset 
write_csv(Clean_Concord_Data,"Clean_Thoreau_Data.csv")


```

```{r}
# Species with no TSN
na_clean <- Clean_Concord_Data[is.na(Clean_Concord_Data$TSN), ]
```
