---
title: "Concord, MA 1951-2020 (plants and birds)"
author: "Karen Ornelas"
date: "2024-03-05"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

load tidyverse
```{r}
install.packages("tidyverse")
library(tidyverse)
```

Object to edit csv file -> CHANGE reference to a folder on desktop
```{r}
Concord_data <- Concord_Phenology_Set_1851_2020
```

Remove all unnescary columns before turning the data into long format 
```{r}
Concord_data <- Concord_data [-c(4,5,6,7,8)]
```


Turn data into long format
```{r}
# change all data into interegers
Concord_data <- Concord_data %>%
  mutate(across(5:125, as.numeric))

Concord_long <- Concord_data %>% 
  pivot_longer(
    cols = 5:125,
    names_to = "year",
    values_to = "DOY")
```


remove na in duration and this will also take out our ice out
```{r}
Concord_long <- 
  Concord_long %>% 
    drop_na()
```


add column for lat and long 
```{r}
Concord_long <- Concord_long %>%
  mutate(Longitude = "-71.3333333")

Concord_long <- Concord_long %>%
  mutate(Latitude = "42.4500000")

```


Create a new test database with only scientific name to match TSN
```{r}
install.packages("taxize")
library(taxize)

Concord_long$Scientific_name <- paste(Concord_long$Genus,Concord_long$Species)

test_df <- distinct(Concord_long,Scientific_name)
test_df <- Concord_long %>% 
  filter(Kingdom == "Plantae") %>% 
  distinct(Scientific_name)
test_df <- test_df %>%
  mutate(TSN = get_tsn(test_df$Scientific_name, rows = 1))


```

Use Fuzzy function in worldflora 
```{r}
install.packages("WorldFlora")
library("WorldFlora")

WFO.download()
WFO.remember("classification.csv")
```

Ellen's attempts to get `WFO.match.fuzzyjoin()` working
```{r}
plants <- Concord_long %>% 
  filter(Kingdom == "Plantae") %>% 
  distinct(Scientific_name)

# use fuzzy match to correct spelling and get current names
test <- plants[1:10,]
fuzzy_match <- WFO.match.fuzzyjoin(spec.data = plants$Scientific_name, WFO.data = WFO.data, fuzzydist.max = 3)

# have WFO condense the above dataframe down to one clear match per species
one_to_one <- WFO.one(fuzzy_match)

# save the output so we don't have to run those again because they take up wild amounts of RAM!
write_csv(one_to_one, "corrected_taxonomy.csv")

# run the updated names through `taxize` to get updates TSN
# only one not found - has incorrect scientfice name "Lysimachia arvensis" should be "Anagallis arvensis" 
one_to_one_TSN <- one_to_one %>% 
  mutate(TSN = get_tsn(one_to_one$scientificName, rows = 1))
```

Pull out NA TSN values 
Trying to determine what the issue is 
```{r}
NA_TSN <- is.na
```


unique values of pheno
```{r}
unique(Concord_long$Phenophase)
```


